sources_summary

### 2025-12-16 manuscript/sources_summary.qmd
lit review

>>>>

# Josie Lit Review

In an early evaluation of tournament design and ranking accuracy, @Carroll1883 evaluated paired comparisons and the method of assigning prizes in Lawn Tennis Tournaments. He argued that the standard single-elimination tournament format often failed to rank competitors correctly, as strong players could be eliminated early due to unfavorable pairings or random variation in match outcomes. To address this, he proposed a system of rules for conducting tournaments that would take less time while producing more equitable results. Carroll also suggested reforms to match scoring, including abolishing the concept of “sets” and instead defining a match as a sequence of individual “games,” with the goal of reducing randomness in outcomes. He concluded that his new methods eliminate the element of luck in tournaments so competitors are fully evaluated on their skills, making it possible to predict who will win.

While Carroll focused on improving tournament rules, Kendall and Smith examined rankings as a measurement problem. They evaluated the method of paired comparisons, where objects are compared only two at a time rather than against all others simultaneously. They argue that the common ranking method of comparison performs poorly when the object quality being measured cannot be represented linearly. Instead, they analyzed the method of paired comparisons, in which an observer expresses a preference between two objects at a time, and a final ranking is inferred after many comparisons are made. Kendall and Smith also explored how these preferences can be represented in tabular or geometric form, emphasizing that inconsistent or “circular” preferences may occur, such as situations in which A is preferred to B, B to C, and C to A. In addition, they examined agreement among multiple judges, distinguishing between disagreement caused by poor judgment and disagreement resulting from indistinguishable differences in object quality. They derive a measurement to quantify the agreement between judges, which is based on Kendall’s Tau.

In The World Series Competition, @Mosteller1952 shifted the focus from designing ranking systems to evaluating how well observed competition outcomes reflect underlying team quality. Rather than assuming that tournament results accurately rank teams, he estimated the probability that the better team wins an individual contest. Using historical data from the World Series, Mosteller modeled game outcomes as events influenced by team strength and random variation. His analysis demonstrated that even in a best-of-seven series, the best team does not always win, highlighting the role of chance in tournament outcomes. His work provided a foundation for evaluating tournament formats by linking design choices, such as the number of games played, to the probability that final outcomes reflect true team ability.

Building on the paired comparison framework, @BradleyTerry1952 introduced a probabilistic model for paired comparisons that became a cornerstone of modern statistical ranking theory. Their model assigns a positive “strength” parameter to each competitor and defines the probability that one competitor beats another as a function of the ratio of their strengths. This formulation allows researchers to infer an overall ranking from observed pairwise outcomes by estimating these strength parameters from data, often by using maximum likelihood methods. The Bradley-Terry model has become widely used in sports analytics and preference modeling because it translates binary outcomes from games or comparisons into a quantitative ranking of competitors.

@Glenn1960-mn returned to the comparison of tournament formats, offering one of the earliest evaluations of tournament "effectiveness." Focusing on competitions with four participants, Glenn analyzed six tournament designs, including round-robin, variations of single-elimination, and double-elimination formats. He compared these designs using two criteria: the probability that the strongest competitor wins and the expected number of games required. By modeling match outcomes probabilistically, Glenn found that tournaments with more games are more accurate but less efficient. His results showed that round-robin tournaments maximize ranking accuracy, while elimination formats prioritize efficiency.
